Dynamical modeling of tidal streams

### 2013-11-20 tex/ms.tex
streams: new section 2

>>>>

\section{The dynamics of tidal streams}\label{sec:dynamics}

The dynamics of tidal streams both in real (position--velocity) and
action--angle space has been described by various previous authors
\citep[\eg,][]{Helmi99a,Tremaine99a,SomeJohnstonPaper,Sanders13a}. I
provide here a brief description of the dynamics of tidal streams that
is relevant for what follows.

In real space, a tidal stream forms as stars are stripped from a
progenitor some time $\Delta t$ in the past and then evolve (largely)
independently in a galaxy's host potential, moving away both in
position and velocity from the progenitor object as time goes
on. Thus, a star was offset by a small amount $(\Delta \vecx,\Delta
\vecv)$ from the position of the progenitor $(\vecx^p,\vecv^p)(-\Delta
t)$ at that time. Then the star orbited under the influence of the
same gravitational potential $\Phi$ as the progenitor until the
present day $(t=0)$, when it was observed: $(\vecx,\vecv)(t=0)$ =
$\vech_{\Delta t}(\vecx^p(-\Delta t)+\Delta \vecx,\vecv^p(-\Delta
t)+\Delta \vecv)$, where $\vech$ denotes the Hamiltonian flow from
$t=-\Delta t$ to $t=0$ (see \citealt{binneytremaine}); this is simply
the orbit of the star between these times. In phase-space this orbit
can be calculated by solving Hamilton's equations: $\dot{\vecx} =
\vecv; \dot{\vecv} = - \dd \Phi / \dd \vecx$.

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\Delta t$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real-space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this offset is responsible for the spreading of the stream
over long stretches on the sky. Because both the source's angles and
the star's angles increase linearly with time, albeit at different
frequencies, the difference in the angles also increases linearly in
time. Therefore at time $t=0$ the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}
As the initial offset $\Delta \veca(-\Delta t$ in angles is small
compared to the growth $\Delta\veco\Delta t$, $\Delta \veca(t=0)
\approx \Delta \veco \Delta t$.

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial^2 H
/ \partial \vecj \partial \vecj) \Delta \vecj$, in the limit of small
action offsets, as $\veco = \partial H / \partial \vecj$. For an
approximately one-dimensional stream to form, the Hessian matrix
$(\partial^2 H / \partial \vecj \partial \vecj)\large|_{\vecj^p}$ has
to be dominated by a single large eigenvalue; in this case the angle
difference of stars stripped from the stream at any time will fall
along approximately the same direction $\unitvec\veco$, and the stream
will essentially be one dimensional. In detail, because the
distribution of initial offsets $\Delta \vecj$ is not isotropic, the
direction $\unitvec \veco$ is not just determined by the potential
(through $H$) and the progenitor orbit, but also by the distribution
of $\Delta \vecj$, which is set through a combination of the internal
dynamics of the progenitor and the stripping process (see discussion
in \citealt{Sanders13a}). For dynamical modeling we can be agnostic
about the relative importance of all of these contributors to
$\unitvec \veco$ by flexible modeling of the distribution of $\Delta
\vecj$.
